A Mechanised Library for Generic (Co)Programming

### 2024-02-28 paper/paper.tex
Intro

As mentioned, some functions, though well-defined, cannot be accepted by the
proof assistant. One example is the quicksort algorith, written in OCaml below:
\begin{minted}{OCaml}
  let divide = function
  |  [] -> None
  |  x::xs -> Some (x, List.partition (fun y -> y < x) xs);;

  let rec quicksort xs = match divide xs with
    | None -> []
    | Some (x, (smaller, larger)) -> quicksort smaller @ (x::quicksort larger)
\end{minted}
While \ocaml{quicksort} is a well-defined mathematical function it cannot be
accepted by a proof assistant.  The reason is that the \ocaml{divide}
function \emph{destructuring} the input ``dives'' deeper in the input returning
two sublists with the head as a pivot.

@@

Despite the benefits of hylomorphisms for reasoning about program equivalences,
not many approaches mechanise structured recursion schemes in proof assistants.
A recent approach by Abreu et al.\~\cite{AbreuDHJMS23} encode an algebraic
approach to divide-and-conquer computations, in which termination is entirely
enforced by the typing discipline of the recursion scheme. However, although
their approach solves the problem of termination proofs, as well as the
performance of the code that is run \emph{within Coq}, it no longer allows the
extraction of clean, naturally-looking code.  By naturally-looking and clean,
we mean code that (1) preserves the recursive structure intended by the
programmer; and (2) does not lead to unsafe casts \ocaml{Obj.magic} in the
generated code.

>>>>

As mentioned, some functions, though well-defined, cannot be accepted by the
proof assistant. One example is the quicksort algorith, written in OCaml below:
\begin{minted}{OCaml}
  let divide = function
    | [] -> None
    | x::xs -> Some (x, List.partition (fun y -> y < x) xs);;
  let rec quicksort xs = match divide xs with
    | None -> []
    | Some (x, (smaller, larger)) -> quicksort smaller @ (x::quicksort larger)
\end{minted}
While \ocaml{quicksort} is a well-defined mathematical function it cannot be
accepted by a proof assistant.  The reason is that the \ocaml{divide}
function \emph{destructuring} the input ``dives'' deeper in the input returning
two sublists with the head as a pivot.

@@

Despite the benefits of hylomorphisms for reasoning about program equivalences,
not many approaches mechanise structured recursion schemes in proof assistants.
A recent approach by Abreu et al.~\cite{AbreuDHJMS23} encode an algebraic
approach to divide-and-conquer computations, in which termination is entirely
enforced by the typing discipline of the recursion scheme. However, although
their approach solves the problem of termination proofs, as well as the
performance of the code that is run \emph{within Coq}, it no longer allows the
extraction of clean, naturally-looking code.  By naturally-looking and clean,
we mean code that (1) preserves the recursive structure intended by the
programmer; and (2) does not lead to unsafe casts \ocaml{Obj.magic} in the
generated code.
